
# Is knee strength change over time between two groups different

## Download and load libraries {-}

```{r, message=FALSE, warning=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, # All purpose wrangling for dataframes
               readr, # importing and writing data
               openxlsx, # writing excel documents
               purrr, # for piping %>%
               knitr,
               moderndive,
               kableExtra,
               infer)

```

## Import data {-}

```{r message=FALSE, warning=FALSE}


subjdat <-  read_delim (file = "data/SubjectData.txt", # file name placed into a folder called data
                        delim = "\t", # file separater
                        col_names = TRUE) # the first row of the data is a header row

scalfac <-  read_delim (file = "data/ScaleFact.txt", # file name
                        delim = "\t", # file separater
                        col_names = TRUE) # the first row of the data is a header row

strn <-  read.xlsx (xlsxFile = "data/STRENGTH.xlsx")

```

## Data manipulation

I need a dataframe which contains the subject identifier, time, group, and knee strength `kexttorq` which is normalized to each participant's body mass.

1. Convert all colnames to lower cases. **Rationale:** Simply to make my life easier.

```{r}

colnames (strn) <-  tolower(colnames (strn))
colnames (subjdat) <-  tolower(colnames (subjdat))

```

2. Check column names for each dataframe. **Rationale:** I want to combined the two dataframes of `strn` and `subjdat`, and I need to make sure that the common columns are identically labelled. 

```{r}
colnames (strn) 
```
```{r}
colnames (subjdat)
```


3. Keep the columns you want by creating two additional objects

```{r}

# I need subj, and body mass from the subjdat dataframe
subjdat.sub <- subjdat %>%
  select (subj, wt)

# I need subj, time, group, side and kexttorq from the strn dataframe
strn.sub <- strn %>%
  select (subj, time, group, side, set, rep, kexttorq)

new.df <- strn.sub %>%
  inner_join(subjdat.sub, by = c("subj"))

head (new.df)
```

4. Divide `kexttorq` by `wt` to get normalized knee strength

```{r}
new.df <- new.df %>%
  mutate (ktorq.norm = kexttorq/wt)
```

5. Because I am interested in the change in `kexttorq.norm` between `PRE` and `POST`, I need to take create two variables ->  `kexttorq.norm.pre` and `kexttorq.norm.post`. To make life simpler, I will create one`kexttorq.norm` value for each subject at each time point.  

```{r}
new.df <- new.df %>%
  group_by (subj, time, group) %>%
  summarize (ktorq.norm = mean (ktorq.norm)) %>%
  ungroup()

head (new.df)
```

6. I will than spread the data to make it wide.

```{r}
new.df <- new.df %>%
  spread (key = time, value = ktorq.norm)

head (new.df)

```

7. I want to create a new variable that is the difference in `kexttorq.norm` which is `POST` strength minus `PRE` strength

```{r}
new.df <- new.df %>%
  mutate (diff = POST - PRE)

head (new.df)
```


7. Let's get some summary statistics per group.**Can you guess** if `diff` is different in group `G` compared to `T`?


```{r }

# grouped summaries
new.df %>%
  group_by(group) %>%
  summarize (Mean = mean (diff, na.rm = T),
             Median = median (diff, na.rm = T),
             Max = max (diff, na.rm = T),
             Min = min (diff, na.rm = T),
             Sd = sd (diff, na.rm = T),
             Iqr = IQR (diff, na.rm = T),
             quant25 = quantile (diff, probs = 0.25, na.rm = T),
             quant75 = quantile (diff, probs = 0.75, na.rm = T))


```

7. Visualize the individual points of the data, so I will create a cleaveland dotplot

```{r}

ggplot(data = new.df) + 
  geom_point(aes(y = seq(1,nrow(new.df),1), x = diff)) +
  ylab ("row number") + # Label as you like
  xlab ("ankle torque") # Label as you like
```

8. Create a bar plot with error bars to see if differences stand out already.


```{r, fig.cap='ankle torque by group', message=FALSE, warning=FALSE}

# generate a summarized dataframe 
df.plot = new.df %>%
  group_by(group) %>%
  summarize (Mean = mean (diff, na.rm = T), 
             Sd = sd (diff, na.rm = T)) %>%
  ungroup()

# Combined plot
ggplot(data = df.plot) + 
geom_bar (aes(y = Mean, x = group), stat = "identity") +
geom_errorbar(aes(ymin=Mean -Sd, ymax=Mean +Sd, x = group), 
            width=.2) +
ylab ("Normalized Knee torque change") +
xlab ("Group")
```

## Hypothesis testing

1. Get the null distribution in a world where two groups have identical change in knee strength

```{r}

null_distribution <-new.df %>%
  specify (formula = diff ~ group)%>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("G", "T"))

head (null_distribution)
```

2. Get the observed difference in change in strength between two groups

```{r}

obs_diff <-new.df %>%
  specify (formula = diff ~ group)%>% 
  calculate(stat = "diff in means", order = c("G", "T"))

obs_diff
```

3. visualize the p-value 


```{r}

visualize(null_distribution, binwidth = 0.05)
```

```{r}

visualize(null_distribution, binwidth = 0.05)+ 
  shade_p_value(obs_stat = obs_diff, direction = "greater")

```

Get the p-value

```{r}

get_p_value(null_distribution, obs_stat = obs_diff, direction = "greater")

```

4. Calculate the mean difference in change in strength between the two groups and its confidence interval.

```{r }

bootstrap_distribution <- new.df %>%
  specify (formula = diff ~ group)%>% 
  generate(reps = 1000, type = "bootstrap") %>% # change this from permute to bootstreap
  calculate(stat = "diff in means", order = c("G", "T"))

```

get the CI

```{r}

percentile_ci <- bootstrap_distribution %>% 
  get_confidence_interval(level = 0.95, type = "percentile")
percentile_ci

```

```{r }

visualize(bootstrap_distribution) + 
  shade_confidence_interval(endpoints = percentile_ci)

```